What is an Ancestral Recombination Graph?

### 2022-02-02 paper.tex
Modified first sentence, added references

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately, however, there is substantial confusion over
terminology and little agreement on shared standards for data interchange,
significantly hampering progress in this important area.
This confusion and lack of agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a stochastic process (the coalescent with recombination) together with its 
representation, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous, fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individual's monoploid genomes, an edge
% jk-note: this is clumsy
defines a genetic ancestry relationship between two nodes,
and annotations on those edges define the precise patterns of
ancestry in intervals along the genome.
We use this framework to define some classifications and metrics of ARG nodes,
and discuss the properties of some recently developed inference methods.
% jk-note: this ends a bit flat - we should say other stuff like we clarify
% how the ARG relates the pedigree, etc, etc. Let's go through again when
% we have a reasonable draft.
\end{abstract}

@@

\section*{Introduction}
Inferring the genealogical relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of non (or
rarely) recombining organisms [refs]. The tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these trees [cite?], and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the
trees~\citep{tajima1983evolutionary,tavare1984line,fu1995statistical}
such as allele frequencies and site frequency spectra~\citep{ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical genetics.

@@

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
\end{enumerate}

@@

Hudson's algorithm operates by tracking the state of a set of ancestral
lineages as we go backwards in time.
% jk-note unclear here about the best choice of notation. We could have
% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
% stick with the original tuple notation for now.
%
% Each lineage consists of a set of
% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
% such that for each $1 \leq j \leq \left| l \right|$,
% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
% $\mathbf{a}_j$ is an integer
% tracking the number of samples the lineage is ancestral to over the interval.
% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
% covered by the lineage, respectively.
Each lineage consists of a list of
disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples to which the lineage is ancestral over that interval.
(We also usually track the tree node associated with each segment, but
that is not important for our purposes here so we omit it.)
If we have $n$ samples and genome of length $m$, the process begins with $n$ lineages
of the form $[(0, m, 1)]$. The process then works backwards in time from
the present day as a series of random common ancestor or recombination events.
Recombination events occur at a rate determined by the amount of ancestral material and
the way in which it is distributed along a chromosome.
Let $L$ be the set of ancestors at a given time $t$. Recombination events
happen at rate $\rho \nu / (m - 1)$ where
\[
\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
\]
is the number of available `links' that may be broken. Thus, the rate of
recombination is determined by the left- and right-most extent of the
ancestral material carried by each lineage. At a recombination
event we choose one of these links uniformly and break it. Given a lineage
$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
$x_1$ and $x_2$ such that FILL IN DETAILS

When $k = |L|$ lineages are present, common ancestor events
occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
are chosen uniformly at random and their ancestry segments merged.
If we have overlapping intervals of ancestry from the two lineages,
say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
\emph{coalescence} occurs and ancestor represented by the current event
will be present as a node (at least) in the marginal trees covering
the interval $[\ell, r)$. The result of this coalescence is a segment
$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
we have found the most recent common ancestor of all samples in
the interval $[\ell, r)$ and so we do not need to simulate its history any further.
Nonoverlapping intervals of ancestry from the two lineages are included
in the resulting lineage without changes. Eventually, as the process continues,
we find resultant lineages in which all segments have fully coalescenced,
and so the number of extant lineages gradually dwindles down to zero.

@@

The state-space of this process is much simpler than Hudson's algorithm, which
greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the ARG back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the complete ancestry of a sample
in Hudson's algorithm is
$O(\rho^2)$~\citep[Eq.~5.10]{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the vast majority of the events that occur in the ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

@@

\section*{The ARG as a data structure}
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}. These methods met with limited success
due to the overwhelmingly large and
horribly structured [Jere: what's the right way to say this?]
state space of ARGs. The focus subsequently shifted to
the more tractable---but still
NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a data structure became less clear at this point,
with \citet{minichiello2006mapping} explicitly arguing for
the data structure interpretation.

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage format.

[Not sure how to make a narrative out of this, but here are the
basic problems:]
\begin{enumerate}
\item The encoding is hamstrung by the interpretation of
nodes as \emph{events} of a particular type. We can either
have common ancestor or recombination events. This is true
in the coalescent, but is not in a more general Wright-Fisher
model.
%% Note - I'm not sure this is true, if the nodes are *events* rather than
%% genomes, unless you want events (rather than genomes) to occur at
%% integer generations.
\item We can also only have one crossover per recombination event.
This is unrealistic because things like gene conversion are common. It also
means we can't have back and forth recombination between
two parents, or represent the transmission of multiple chromosomes.
\item The passage of ancestral material along edges requires
extrinsic rules which must be shared by all implementations.
These rules can be quite restrictive when generating, and
in practise some have found it simpler to add explicit
annotations to the edges.
\item Generating local trees is inefficient, essentially requiring
$O(n T)$ time (or more?)
\end{enumerate}

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
\caption{\label{fig-recombination-nodes}
Recombination nodes are optional in an edge-annotated ARG. (a) A simulated
ARG including all recombination nodes (red). (b) The same ARG having removed both
recombination nodes and nodes of common ancestors in which no genetic
lineages coalesce (cyan). (c) The sequence of trees corresponding to (b): if we were
to plot out the trees corresponding to (a) we would have more trees, but the topologies
and MRCA nodes at each genomic location would be identical. For clarity, nodes are
plotted by ranked time rather than true time on the y-axis.
% HYW: note that we also truncate nodes for portions in which they are unary,
% so that nodes which are coalescent nodes over only some of their length
% (blue-green) become coalescent nodes over their entire length (green). This
% can, however, throw away some useful information, for example node 4 has
% only a single parent edge in (a), but 2 parent edges (one of which appears
% to bypass node 11) in (b). In fact, this edge should go through node 11, but
% there may be no way to detect this from real data.
}
\end{figure}

>>>>

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately, however, there is substantial confusion over
terminology and little agreement on shared standards for data interchange,
significantly hampering progress in this important area.
This confusion and lack of agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a representation of the coalescent with recombination
stochastic process as a graph, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous, fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individuals monoploid genomes, an edge
% jk-note: this is clumsy
defines a genetic ancestry relationship between two nodes,
and annotations on those edges define the precise patterns of
ancestry in intervals along the genome.
We use this framework to define some classifications and metrics of ARG nodes,
and discuss the properties of some recently developed inference methods.
% jk-note: this ends a bit flat - we should say other stuff like we clarify
% how the ARG relates the pedigree, etc, etc. Let's go through again when
% we have a reasonable draft.
\end{abstract}

@@

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}. The tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these trees [cite?], and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the trees such as allele
% TODO cite the other papers that preceded ralph2020efficiently here.
frequencies~\citep{ralph2020efficiently}, site frequency spectra, patterns of linkage
disequilibrium~\citep{mcvean2002genealogical} and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical genetics.

@@

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a diploid pedigree rather than
the coalescent approximations and spatial annotations are attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
\end{enumerate}

@@

Hudson's algorithm operates by tracking the state of a set of ancestral
lineages as we go backwards in time.
% jk-note unclear here about the best choice of notation. We could have
% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
% stick with the original tuple notation for now.
%
% Each lineage consists of a set of
% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
% such that for each $1 \leq j \leq \left| l \right|$,
% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
% $\mathbf{a}_j$ is an integer
% tracking the number of samples the lineage is ancestral to over the interval.
% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
% covered by the lineage, respectively.
Each lineage consists of a list of
disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples the lineage is ancestral to over the interval.
(We also usually track the tree node associated with each segment, but
that is not important for our purposes here so we omit it.)
If we have $n$ samples and genome of length $m$, the process begins with $n$ lineages
of the form $[(0, m, 1)]$. The process then works backwards in time from
the present day as a series of random common ancestor or recombination events.
Recombination events occur at a rate determined by the amount and spatial
distribution of ancestral material among lineages.
Let $L$ be the set of ancestors at a given time $t$. Recombination events
happen at rate $\rho \nu / (m - 1)$ where
\[
\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
\]
is the number of available `links' that may be broken. Thus, the rate of
recombination is determined by the left- and right-most extent of the
ancestral material carried by each lineage. At a recombination
event we choose one of these links uniformly and break it. Given a lineage
$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
$x_1$ and $x_2$ such that FILL IN DETAILS

When $k = |L|$ lineages are present, common ancestor events
occur at rate $\binom{k}{2}$. In a common ancestor event two lineages
are chosed uniformly at random and their ancestry segments merged.
If we have overlapping intervals of ancestry from the two lineages,
say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
\emph{coalescence} occurs and ancestor represented by the current event
will be present as a node (at least) in the marginal trees covering
the interval $[\ell, r)$. The result of this coalescence is a segment
$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
we have found the most recent common ancestor of all samples in
the interval $[\ell, r)$ and so we do not need to simulate its history any further.
Nonoverlapping intervals of ancestry from the two lineages are included
in the resulting lineage without changes. Eventually, as the process continues,
we find resultant lineages in which all segments have fully coalescenced,
and so the number of extant lineages gradually dwindles down to zero.

@@

The state-space of the ARG is much simpler than Hudson's algorithm, which
greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the ARG back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the complete ancestry of a sample
in Hudson's algorithm is
$O(\rho^2)$~\citep[Eq.~5.10]{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the vast majority of the events that occur in the ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

@@

\section*{The ARG as a data structure}
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}. These methods met with limited success
due to the overwhelmingly large and
horribly structured [Jere: what's the right way to say this?]
state space of ARGs. The focus subsequently shifted to
the more tractable---but still
NP-hard~\citep{wang2001perfect}---problem of computing
the miniumum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a data structure became less clear at this point,
with \citet{minichiello2006mapping} explicitly arguing for
the data structure interpretation.

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination, but
it has significant drawbacks as the basis for a general purpose
storage format.

[Not sure how to make a narrative out of this, but here are the
basic problems:]
\begin{enumerate}
\item The encoding is hamstrung by the interpretation of
nodes as \emph{events} of a particular type. We can either
have common ancestor or recombination events. This is true
in the coalescent, but is not in a more general Wright-Fisher
model.
\item We can also only have one crossover per recombination event.
This is unrealistic because things like gene conversion are common,
and means we can't have back and forth recombination between
two parents.
\item The passage of ancestral material along edges requires
extrinsic rules which must be shared by all implementations.
These rules can be quite restrictive when generating, and
in practise some have found it simpler to add explicit
annotations to the edges.
\item Generating local trees is inefficient, essentially requiring
$O(n T)$ time (or more?)
\end{enumerate}

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
\caption{\label{fig-recombinati on-nodes}
Recombination nodes are optional in an edge-annotated ARG. (a) A simulated
ARG including all recombination nodes (red). (b) The same ARG having removed both
recombination nodes and nodes of common ancestors in which no genetic
lineages coalesce (cyan). (c) The sequence of trees corresponding to (b): if we were
to plot out the trees corresponding to (a) we would have more trees, but the topologies
and MRCA nodes at each genomic location would be identical. For clarity, nodes are
plotted by ranked time rather than true time on the y-axis.
% HYW: note that we also truncate nodes for portions in which they are unary,
% so that nodes which are coalescent nodes over only some of their length
% (blue-green) become coalescent nodes over their entire length (green). This
% can, however, throw away some useful information, for example node 4 has
% only a single parent edge in (a), but 2 parent edges (one of which appears
% to bypass node 11) in (b). In fact, this edge should go through node 11, but
% there may be no way to detect this from real data.
}
\end{figure}
